---
title: "Exploratory Data Analysis - Homework 5"
author: "Ganesh Nagarajan, gnagaraj@indiana.edu"
date: "October 30, 2015"
output: pdf_document
---

1. Median Polish Column first

```{r}
#Input the columns
c1<-c(25.3,32.1,38.8,25.4)
c2<-c(25.3,29.1,31.0,21.1)
c3<-c(18.2,18.8,19.3,20.3)
c4<-c(18.3,24.3,15.7,24.0)
c5<-c(16.3,19.0,16.8,17.5)
mat<-cbind(c1,c2,c3,c4,c5)

twoway.median2 <- function(mat){ # first column then row
  meff.MP <- median(mat)
  beff.MP <- apply(mat,2,median,na.rm=T)  # column medians
  mat.res <- mat - matrix(rep(beff.MP,each=nrow(mat)),byrow=F,nrow=nrow(mat));
  aeff.MP <- apply(mat.res,1,median,na.rm=T) # row effect
  beff.MP <- beff.MP - median(beff.MP)  # column effect
  res.MP <- mat.res - matrix(rep(aeff.MP,each=ncol(mat)),byrow=T,ncol=ncol(mat))
  list(overall=meff.MP, row=aeff.MP, col=beff.MP, res=res.MP)
}

lv1<-twoway.median2(mat)
lv1
lv2<-twoway.median2(lv1$res)
lv2
overallRow<-lv1$row+lv2$row
print(overallRow)
overallCol<-lv1$col+lv2$col
print(overallCol)
overall<-lv1$overall+lv2$overall
print(overall)
stem(c(lv2$res),2)
```

It can be seen that after two iterations, both the row first and column first median polish results in not extremely different, but similar overall and row/column effects.

2.Personal Consumption Expenditures Dataset

```{r}
diag.MP <- function(fit){
  fit.comp <- matrix(fit$row,ncol=1) %*% matrix(fit$col,nrow=1)/fit$overall
  plot(fit.comp, fit$res,xlab="Comparison value",ylab="Residual",cex=0.5)
  abline(v=0,h=0,lty=2)
  ls <- lm(c(fit$res)~c(fit.comp))
  abline(ls,col="red",lty=3) 
  rr <- run.rrline(fit.comp,fit$res,iter=10) 
  abline(rr$a, rr$b, col="red")
  pwr1 <- 1 - rr$b
  pwr2 <- 1 - ls$coef[2]
  title("",paste("Approximate power =",format(round(pwr1,2))," or ", format(round(pwr2,2))))
}

symbolPlot<-function(mat){
  result<-medpolish(mat)
  res<-c(result$residuals)
  genNos<-expand.grid(1:5,1:5)
  plotvar<-cbind(genNos$Var2,genNos$Var1,res)
  pos<-plotvar[plotvar[,3]>=0,]
  max<-sum(abs(pos[,3]))
  symbols(pos[,1],pos[,2],squares = 0.2*(abs(pos[,3]/(max))),inches = FALSE,xlab="Columns",ylab="Rows",main="Symbol Plot")
  pos<-plotvar[plotvar[,3]<0,]
  symbols(pos[,1],pos[,2],circles = 0.2*(abs(pos[,3]/(max))),inches = FALSE,add = TRUE)
}

rrline1 <- function(x,y) {
  n3 <- floor((length(x)+1.99)/3)
  x.order <- order(x)
  medxL <- median(x[x.order][1:n3])
  medxR <- median(rev(x[x.order])[1:n3])
  medyL <- median(y[x.order][1:n3])
  medyR <- median(rev(y[x.order])[1:n3])
  slope1 <- (medyR - medyL)/(medxR - medxL)
  int1 <- median(y - slope1 * x)
  # print(c(paste("Intercept = ", format(round(int1,5))),
  #   paste("Slope = ",format(round(slope1,5)))))
  newy <- y - slope1*x - int1
  sumres <- sum(abs(newy))
  list(a=int1, b=slope1, sumres = sumres, res=newy)
}

run.rrline <- function(x,y,iter=5) {
  out.coef <- matrix(0,iter,3)
  newy <- y
  for (i in 1:iter) {
    rr <- rrline1(x,newy)
    out.coef[i,] <- c(rr$a,rr$b,rr$sumres)
    newy <- rr$res
  }
  dimnames(out.coef) <- list(format(1:iter),c("a","b","|res|"))
  aa <- sum(out.coef[,1])
  bb <- sum(out.coef[,2])
  cc <- sum(abs(y - aa - bb*x))
  res <- y - aa - bb*x
  out.coef <- rbind(out.coef,c(aa,bb,cc))
  print(round(out.coef,5))
  list(a = aa, b = bb, res = res, coef=out.coef)
}

rrline2 <- function(x,y) {
  n <- length(x)
  n3 <- floor((length(x)+1.99)/3)
  x.order <- order(x)
  medxL <- median(x[x.order][1:n3])
  medxR <- median(rev(x[x.order])[1:n3])
  medyL <- median(y[x.order][1:n3])
  medyR <- median(rev(y[x.order])[1:n3])
  medxM <- median(x[x.order][(n3+1):(n-n3)])
  medyM <- median(y[x.order][(n3+1):(n-n3)])
  slope1 <- (medyR - medyL)/(medxR - medxL)
  int1 <- median(y - slope1 * x)
  int2 <- mean(c(medyL,medyM,medyR) - slope1*c(medxL,medxM,medxR))
  newy <- y - slope1*x - int1
  sumres <- sum(abs(newy))
  newy2 <- y - slope1*x - int2
  sumres2 <- sum(abs(newy2))
  list(a=int1, a2=int2, b=slope1, sumres = sumres, sumres2=sumres2, res=newy2)
}

run.rrline2 <- function(x,y,iter=5) {
  out.coef <- matrix(0,iter,3)
  newy <- y
  for (i in 1:iter) {
    rr <- rrline2(x,newy)
    out.coef[i,] <- c(rr$a2,rr$b,rr$sumres2)
    newy <- rr$res
  }
  dimnames(out.coef) <- list(format(1:iter),c("a","b","|res|"))
  aa <- sum(out.coef[,1])
  bb <- sum(out.coef[,2])
  cc <- sum(abs(y - aa - bb*x))
  res <- y - aa - bb*x
  out.coef <- rbind(out.coef,c(aa,bb,cc))
  print(round(out.coef,5))
  list(a = aa, b = bb, res = res, coef=out.coef)
}
# Input Vector
C1<-c(22.2,44.5,59.6,73.2,86.8)
C2<-c(10.5,15.5,29.0,36.5,46.2)
C3<-c(3.53,5.76,9.71,14.0,21.1)
C4<-c(1.04,1.98,2.45,3.40,5.40)
C5<-c(.641,.974,1.80,2.60,3.64)
mat<-cbind(C1,C2,C3,C4,C4)
result<-medpolish(mat)
result
```
Analog R square calcuation
```{R}
Analog_R_Square<- 1-((sum(abs(result$residuals))) /(sum(abs(mat-result$overall))))
Analog_R_Square
```
b) Symbol Plot. The positive are given square, negatives are given circle.
Yes there are patterns. Main diagonal elements have positive residuals, other side has negative elements.
```{r}
symbolPlot(mat)
```
c) Diagnostic plot
It can be noted the residuals are not along the zero axis. Hence must be transformed.
```{r}
plot(result)
diag.MP(result)
```
d) According to the Analog R2, log transform is suggested. Following is the log transform
```{R}
mat<-log(mat)
result<-medpolish(mat)
result
symbolPlot(mat)
```
It can be clearly seen that after transformation, most of the residuals are along zeroes,
hence is a good fit.
```{R}
diag.MP(result)

```
e) Forget it plot - Before and After transformation:
```{R}
forgetitplot <- function(outmpol,outlim=0,...) {
  # outmpol is output of medpolish in library(eda) or library(stats)
  # be sure to assign dimnames to matrix being polished
  oldpar <- par()
  par(fig=c(0,.7,0,1))
  nc <- length(outmpol$col)
  nr <- length(outmpol$row)
  a <- rep(outmpol$row,nc)
 b <- rep(outmpol$col,rep(nr,nc)) 
  sqrt2 <- sqrt(2)
  ab <- cbind((a-b)/sqrt2,(a+b)/sqrt2)
  xrange <- range(ab[,1]) + c(-.1,.1)*(max(ab[,1])-min(ab[,1]))
  yrange <- range(ab[,2]) + c(-.1,.1)*(max(ab[,2])-min(ab[,2]))
  dx <- (xrange[2]-xrange[1])/50
  dy <- (yrange[2]-yrange[1])/50
  plot(ab[,1],ab[,2],axes=F,xlim=xrange,ylim=yrange,xlab="",ylab="",...)
  segments((min(a)-outmpol$col)/sqrt2, (min(a)+outmpol$col)/sqrt2,
           (max(a)-outmpol$col)/sqrt2, (max(a)+outmpol$col)/sqrt2,lty=3)
  segments((outmpol$row-min(b))/sqrt2, (outmpol$row+min(b))/sqrt2,
           (outmpol$row-max(b))/sqrt2, (outmpol$row+max(b))/sqrt2,lty=3)
  # segments((outmpol$row)/sqrt2-min(b), (outmpol$row)/sqrt2+min(b),
  #        (outmpol$row)/sqrt2-max(b), (outmpol$row)/sqrt2+max(b),lty=3)
  yrowloc <-  rep(max(b),nr)
  xrowloc <-  outmpol$row
  # text((xrowloc-yrowloc)/sqrt2-dx,dy+(xrowloc+yrowloc)/sqrt2,format(1:nr))
  text((xrowloc-yrowloc)/sqrt2-dx,dy+(xrowloc+yrowloc)/sqrt2,
       names(sort(outmpol$row)))
  xcolloc <- rep(max(a),nc)
  ycolloc <- outmpol$col
  # text(dx+(xcolloc-ycolloc)/sqrt2,dy+(xcolloc+ycolloc)/sqrt2,format(1:nc))
  text(dx+(xcolloc-ycolloc)/sqrt2,dy+(xcolloc+ycolloc)/sqrt2,
       names(sort(outmpol$col)))
  ynames <- format(round(outmpol$overall + sqrt2*pretty(ab[,2])))
  axis(2,at=pretty(ab[,2]),labels=ynames)
  # add vertical lines when there is an outlier
  if(abs(outlim) > 1e-4) {
    out.index <- which(abs(outmpol$res) > outlim, arr.ind=T)
    # find (r,c) for outlier indices
    zz.x <- outmpol$row[out.index[,1]]
    zz.y <- outmpol$col[out.index[,2]]
    # outlier points at (zz.x-zz.y)/sqrt2, (zz.x+zz.y)/sqrt2
    # draw segment from here to end of residual
    segments((zz.x-zz.y)/sqrt2, (zz.x+zz.y)/sqrt2,
             (zz.x-zz.y)/sqrt2, (zz.x+zz.y)/sqrt2 + outmpol$res[out.index])
  }
  par <- oldpar
  invisible()
}
C1<-c(22.2,44.5,59.6,73.2,86.8)
C2<-c(10.5,15.5,29.0,36.5,46.2)
C3<-c(3.53,5.76,9.71,14.0,21.1)
C4<-c(1.04,1.98,2.45,3.40,5.40)
C5<-c(.641,.974,1.80,2.60,3.64)
mat<-cbind(C1,C2,C3,C4,C4)
rownames(mat)<-c("Food","Household","Medical","Personal","Edication")
colnames(mat)<-c(1940,1945,1950,1955,1960)
result<-medpolish(mat)
forgetitplot(result)

```

It can be seen that since year has lot of elongations and more distributed than type.
Hence year/time has larger effect on the data.

After Transformation
```{R}
mat<-log(mat)
result<-medpolish(mat)
forgetitplot(result)
```
It can be clear seen that after the transformation, the row and column effects are now spreaded than evenly than before the transformation. Hence gives a uniform effect rather than particular points influencing the median polish.

3.Creating the simulated dataset
```{r}
set.seed(1234)
#ti generator
ti<-function(t){((2*t)-1)/100}
t<-sapply(1:50,ti)

#mu-i generator
mui<-function(t){t+0.5*exp(-50*(t-0.5)^2)}
mu<-sapply(t,mui)

#Random Noise
e<-rnorm(50,0,0.5)

#Total function y
y<-(mu*t)+e

#bind them
testDS<-as.data.frame(cbind(t,y))

#Meeting with Dr.king,
u = expression(t + 0.5*exp(-50*t^2 - 12.5 + 50*t))

uder2 = D(D(u, 't'), 't')
ff=function(t){eval({t=t;uder2})}
gg=function(t){ff(t)^2}
j = unlist(integrate(gg,0,1))
j = j[[1]]
r = 1/(2*sqrt(pi))
l = 50^(-1/5)*((0.25*r)/(j*1))^(1/5)

#Print the computed l
print(l)
```
The above is the estimated lambda optimal.

$$
\tfrac{\mathrm{d}^2}{\mathrm{d}t^2}\left[t+0.5{\mathrm{e}}^{{-50{\left(t{-0.5}\right)}^{2}}}\right] = \left(5000{t}^{2}{-5000t}+1200\right){\mathrm{e}}^{{-\frac{100{t}^{2}{-100t}+25}{2}}} $$
$$
\displaystyle\int\limits^{1}_{0} \left(5000{t}^{2}{-5000t}+1200\right){\mathrm{e}}^{{-\frac{100{t}^{2}{-100t}+25}{2}}}\left(5000{t}^{2}{-5000t}+1200\right){\mathrm{e}}^{{-\frac{100{t}^{2}{-100t}+25}{2}}}\,\mathrm{d}t =$$
$$\dfrac{\sqrt{{\pi}}\left(288000\operatorname{erf}\left(5\right){-287625}\right){-500\operatorname{\Gamma}\left(\frac{5}{2},25\right)}+1000\operatorname{\Gamma}\left(\frac{3}{2},25\right)+287500\operatorname{\Gamma}\left(\frac{1}{2},25\right)}{2} = 332.33
$$
```{R}
a = ksmooth(t, y, kernel = "normal", l)
b = ksmooth(t, y, kernel = "normal", 0.4)
c = ksmooth(t, y, kernel = "normal", 0.8)
d = ksmooth(t, y, kernel = "normal", 0.1)
plot(t, y, type = "l", col = "red")
lines(a$x, a$y, col = "green", type = "l")
lines(b$x, b$y, col = "blue", type = "l")
lines(c$x, c$y, col = "orange", type = "l")
lines(d$x, d$y, col = "black", type = "l")
```
The green line is the actual value that was lambda optimal. The it is observed that when the lambda is less, the line seems to be curvy, however a higher lambda straightens the line.

Acknowledgements
1. Online integration and differential calculator http://www.derivative-calculator.net/ and http://www.integral-calculator.com/
